---
title: "Dynamic Foraging Patterns in Antarctic Humpbacks"
author: "Ben Weinstein"
date: "`r Sys.time()`"
output: 
  html_document:
    keep_md: yes
    number_sections: yes
    theme: spacelab
    toc: yes
---

```{r,warning=FALSE,message=FALSE,echo=FALSE}
library(knitr)
opts_chunk$set(echo=F,warning=F,message=F,fig.width = 11,fig.height = 5,cache=F)
library(boot)
library(raster)
library(tidyr)
library(ggplot2)
library(MASS)
library(ggmap)
library(dplyr)
library(chron)
library(gridExtra)
library(stringr)
library(R2jags)
library(maptools)
library(reshape2)
#New model to be run, flag turned off if just updating.
newModel<-T
```

```{r,eval=F}
#Load in data
load("Empirical.RData")
#jags$recompile()
newModel<-F
```

```{r}

#load elevation layer
elev<-raster("InputData/ETOPO1_Ice_g_geotiff.tif")


#read data
mdat<-read.csv("InputData/Antarctic Humpback overlap with krill fisheries -3085052751565071394.csv")
#standardize column names to match the simulation
#Create an animal tag.
mxy <- as(mdat, "data.frame")
mxy$Animal<-mxy$individual.local.identifier
mxy$x<-mxy$location.long
mxy$y<-mxy$location.lat

#remove minke whales, they should already be gone.
minke<-c("131117","131118","131120","154184")
mxy<-mxy[!mxy$individual.local.identifier %in% minke,]

#grab set of animals
mxy<-mxy[mxy$Animal %in% c("112699","121207","131151"),]

#empty coordinates
mxy<-mxy[!is.na(mxy$x),]

#crop by extent
d<-SpatialPointsDataFrame(cbind(mxy$x,mxy$y),data=mxy,proj4string=CRS("+proj=longlat +datum=WGS84"))

d<-crop(d,y=extent(c(-75,-50,-70,-57.5)))

mxy<-as.data.frame(d)

#set datestamp
mxy$timestamp<-as.POSIXct(mxy$timestamp,format="%Y-%m-%d %H:%M:%S.000")

#month and year columns
mxy$Month<-months(mxy$timestamp)
mxy$Year<-years(mxy$timestamp)

#Only austral sping and summer
mxy<-mxy[mxy$Month %in% month.name[1:7],]

#remove empty timestamps
mxy<-mxy[!is.na(mxy$timestamp),]

#remove duplicates
mxy<-mxy[!duplicated(data.frame(mxy$timestamp,mxy$Animal)),]

```

```{r}
#hold copy of orig
oxy<-mxy

#remake spatial object
d<-SpatialPointsDataFrame(cbind(mxy$x,mxy$y),data=mxy,proj4string=CRS("+proj=longlat +datum=WGS84"))

#Plot
pp<-c(mean(mxy$x),mean(mxy$y))
m <- get_map(location=pp,source="google",zoom=4,scale=2,maptype="satellite",color = "bw")
ggmap(m)+geom_path(data=mxy, aes(x=x, y=y,col=as.factor(Animal)),size=.5) + scale_color_discrete(guide="none") + theme_inset()
```

#Correlated random walk

*Process Model*

$$ d_{t} \sim T*d_{t-1} + Normal(0,\Sigma)$$
$$ x_t = x_{t-1} + d_{t} $$

## Parameters

For each individual:

$$\theta = \text{Mean turning angle}$$
$$\gamma = \text{Move persistence} $$

For both behaviors process variance is:
$$ \sigma_{latitude} = 0.1$$
$$ \sigma_{longitude} = 0.1$$

##Behavioral States

$$ \text{For each individual i}$$
$$ Behavior_1 = \text{traveling}$$
$$ Behavior_2 = \text{foraging}$$

$$ \alpha_{i,1,1} = \text{Probability of remaining traveling when traveling}$$
$$\alpha_{i,2,1} = \text{Probability of switching from Foraging to traveling}$$

$$\begin{matrix}
  \alpha_{i,1,1} & 1-\alpha_{i,1,1} \\
  \alpha_{i,2,1} & 1-\alpha_{i,2,1} \\
\end{matrix}$$

With the probability of switching states:

$$logit(\phi_{traveling}) = \alpha_{Behavior_{t-1}}$$

$$\phi_{foraging} = 1 - \phi_{traveling} $$

##Continious tracks

The transmitter will often go dark for 10 to 12 hours, due to weather, right in the middle of an otherwise good track. The model requires regular intervals to estimate the turning angles and temporal autocorrelation. As a track hits one of these walls, call it the end of a track, and begin a new track once the weather improves. We can remove any micro-tracks that are less than three days.
Specify a duration, calculate the number of tracks and the number of removed points. Iteratively.

```{r}
##Time is the beginning of the first point.
step_length=12

sxy<-split(mxy,mxy$Animal)

#time diff function
timed<-function(d,step_length){
  d$j[1]<-0
  for (x in 2:nrow(d)){
    d$j[x]<-as.numeric(difftime(as.POSIXct(d$timestamp[x]),as.POSIXct(d$timestamp[x-1]),units="mins"))/(step_length*60)
  }
  
  #Split out track endings
  ends<-c(1,which(d$j>1),nrow(d))

  for(w in 2:length(ends)){
    d[ends[w-1]:ends[w],"Track"]<-w-1
  }
  
  #remove tracks that are shorter than three days
  track_time<-d %>% group_by(Track) %>% summarize(mt=difftime(max(as.POSIXct(timestamp)),min(as.POSIXct(timestamp)),units="hours")) %>% filter(mt>=24) %>% .$Track
  
  d<-d[d$Track %in% track_time,]
  
  #renumber the tracks
  d$Track<-as.numeric(as.factor(d$Track))
  return(d)
  }

sxy<-lapply(sxy,timed,step_length=12)

#Format matrices for jags
mxy<-rbind_all(sxy)

######recode whales
#mxy$Animal<-as.numeric(as.factor(mxy$Animal))

sxy<-split(mxy,list(mxy$Animal,mxy$Track),drop=TRUE)

sxy<-lapply(sxy,function(x){
#How many observations in each step length segment
x$step<-as.numeric(cut(as.POSIXct(x$timestamp),"12 hours"))
return(x)
})

mxy<-rbind_all(sxy)

#refactor animal
mxy$Animal<-as.numeric(as.factor(mxy$Animal))
```

```{r}
#total number of steps per track/animal
steps_all<-mxy %>% group_by(Animal,Track) %>% summarize(n=length(unique(step)))

# give each step a label
mxy<-mxy %>% group_by(Animal,Track,step) %>% mutate(jStep=1:n())

#Cast time array
j<-acast(mxy,Animal~Track~step~jStep,value.var="j")

#how many observations per individual in each step
mxy$step<-factor(mxy$step,levels=1:max(steps_all$n))
idx<-melt(table(mxy$Animal,mxy$Track,mxy$step))
colnames(idx)<-c("Animal","Track","step","jStep")
idx<-acast(data=idx,Animal~Track~step)

#month array
mxy$MonthF<-as.numeric(factor(mxy$Month,levels=month.name))

MonthA<-acast(mxy,Animal~Track~step,value.var="MonthF",fun.aggregate = min)
MonthA[!is.finite(MonthA)]<-NA

#Individuals
ind=length(unique(mxy$Animal))

#tracks per indivudal
tracks<-mxy %>% group_by(Animal) %>% summarize(tracks=length(unique(Track))) %>% .$tracks

#steps per track
steps<-acast(steps_all,Animal~Track,value.var="n")

#index array for merging
mxy$Index<-1:nrow(mxy)

#obs array
obs<-melt(mxy,measure.vars=c("x","y"))
obs<-acast(obs,Animal~Track~step~jStep~variable)
```

#Create presence map

```{r}
d2<-SpatialPointsDataFrame(cbind(mxy$x,mxy$y),data=data.frame(mxy),proj4string=CRS("+proj=longlat +datum=WGS84"))
#add a dummy code
d2$dummy<-1

r<-raster(d2,ncol=100,nrow=100)
pres<-rasterize(d2,r,field="dummy",fun="count")
plot(pres)

#make pres absence 
plot(ab<-pres>0)
```

## Draw random background points for psuedo-absences

```{r}
#mask out land
elev[elev>0]<-NA

#resample to the observed raster
re<-resample(elev,ab)

#mask out observe locations
temp<-ab

re2<-mask(re,temp,inverse=T)

#take twice as many absence as presence points
psuedo<-sampleRandom(re2,size=sum(ab[],na.rm=T),sp=T,xy=T,na.rm=T)

#create a dataframe of species and absence points
predf<-data.frame(rasterToPoints(ab))
absdf<-data.frame(psuedo) %>% select(x,y) %>% mutate(layer=0)
occ<-rbind_all(list(predf,absdf))

#turn back to spatial points
occ<-SpatialPointsDataFrame(cbind(occ$x,occ$y),data=occ)

#update the raster layer
temp<-ab
plot(presab<-rasterize(occ,ab,field='layer'),main="Presence Absence Raster",col=topo.colors(10),xaxt='n',yaxt='n')
```

#Extract environment

```{r}
#bathymetry
occ$bat<-raster::extract(elev,occ)

#distance to coast

#this is ugly but for the moment make NA 0's, it comes from the aggregating spatial resolutions.
occ$bat[is.na(occ$bat)]<-0
```

#Behavioral Layers

```{r,child="Bayesian/MultiSpecies.R",eval=T}
```

```{r,eval=T}
#source jags file
source("Bayesian/MultiSpecies.R")

#prior cov shape
R <- diag(c(1,1))
data=list(argos=obs,steps=steps,R=R,ind=ind,j=j,idx=idx,tracks=tracks)

#paramters to track
pt<-c("theta","gamma","alpha_mu","state")

if(newModel){
  system.time(jagM<-jags.parallel(model.file = "Bayesian/Multi_RW.jags",data=data,n.chains=2,parameters.to.save=pt,n.iter=10,n.burnin=0,n.thin=1,DIC=FALSE))
}

```


```{r,eval=F}
#update function
recompile(jagM)
jagM <- update(jagM,n.iter=300,n.burnin=0,n.thin=0) 
```

##Chains
```{r,eval=T}

#delete jags objects
#rm(data)
#rm(argos)
#rm(obs)
#rm(j)
gc()

#bind chains
pc<-melt(jagM$BUGSoutput$sims.array)

rm(jagM)
gc()

colnames(pc)<-c("Draw","chain","par","value")

#extract parameter name
pc$parameter<-data.frame(str_match(pc$par,"(\\w+)"))[,-1]

#Extract index
splitpc<-split(pc,pc$parameter)

#single index
splitpc[c("theta")]<-lapply(splitpc[c("theta")],function(x){
    sv<-data.frame(str_match(x$par,"(\\w+)\\[(\\d+)]"))[,3]
    pc<-data.frame(x,Behavior=sv)
    return(pc)
})

##Double index
#single index
splitpc[c("alpha_mu","gamma")]<-lapply(splitpc[c("alpha_mu","gamma")],function(x){
    sv<-data.frame(str_match(x$par,"(\\w+)\\[(\\d+),(\\d+)]"))[,3:4]
    colnames(sv)<-c("Behavior","Month")
    pc<-data.frame(x,sv)
    return(pc)
})

#State index
splitpc[c("state")]<-lapply(splitpc[c("state")],function(x){
#As matrices
    sv<-data.frame(str_match(x$par,"(\\w+)\\[(\\d+),(\\d+),(\\d+)]"))[,3:5]
    colnames(sv)<-c("Animal","Track","step")
    pc<-data.frame(x,sv)
})

#bind all matrices back together
pc<-rbind_all(splitpc)
rm(splitpc)

#make behavior a factor
pc$Behavior<-as.factor(pc$Behavior)
```

```{r,fig.height=10}
#plot all but phi and state
ggplot(pc[!pc$parameter %in% c("phi","state"),],aes(x=Draw,y=value)) + facet_wrap(~par,scales="free",ncol=3) + geom_line(aes(col=as.factor(chain))) + labs(col="Chain")
```

### View behavioral model posteriors

```{r,fig.height=20}
#Plot
ggplot(data=pc[!pc$parameter %in% c("phi","state"),],aes(x=value)) + geom_histogram() + facet_wrap(~par,scales='free',ncol=3) + theme_bw()  + ggtitle("Estimated parameters")
```

##Behavioral Prediction

```{r}
#mode state for each position
mode_state<-pc %>% filter(pc$parameter %in% "state") %>% group_by(Animal,Track,step) %>% summarize(state=as.numeric(names(which.max(table(value)))))

mxy<-merge(mxy,mode_state,by=c("step","Track","Animal"))

#ensure order for plotting
mxy<-mxy %>% arrange(Animal,Track,step,jStep)
```

#Foraging suitabality

Fow now, a cell gets labeled as suitable for foraging if it ever has predicted foraging state.

```{r}
d3<-SpatialPointsDataFrame(cbind(mxy$x,mxy$y),data=data.frame(mxy),proj4string=CRS("+proj=longlat +datum=WGS84"))
#add a dummy code
d2$dummy<-1

forage<-rasterize(d3,r,field="state",fun="max")
plot(forage,col=topo.colors(10))
```

```{r}
#combine into one data.frame, need to ensure spatial structure lines up
foragedf<-rasterToPoints(forage)
colnames(foragedf)[3]<-"Behavior"

#merge with occurrence data.
k<-merge(occ,foragedf,by=c("x","y"))

#ensure that no behavior in absence points.
table(k$layer,k$Behavior)
```

```{r}
save.image("Empirical.RData")
```